Qualifying Report

### 2019-01-29 Source/QualifyingReport/QualifyingReport.tex
QR

In 2009, the Department of Health designed it's National Dementia strategy and as part of this made early diagnosis and support one of it's key priorities \cite{England2009}. A lot of work has gone into trying to find ways of improving the early diagnosis of Alzheimer's Disease (AD) and Mild Cognitive Impairment (MCI) with research focused on two areas - identifying biological markers and analyzing the cognitive decline of those who are suspected to have the disease\cite{Taler2008}. As described above, the numbers of those suffering from AD and MCI are going to increase as the population ages \cite{Prince2015} and thus it is important that we utilize technology wherever possible to aid clinicians in the detection of MCI and AD. At the present time diagnosis is typically conducted at memory clinics by trained clinicians\cite{Boschi2017}. I theorize that we may be able to enable an earlier diagnosis of those with MCI and AD using samples of spontaneous speech, natural language processing (NLP) and machine learning (ML).

@@

There is a large body of research that looks at the decline in language in those with MCI and AD\cite{Taler2008, Boschi2017}. However there is conflicting evidence in these studies about which declining language factors are associated of MCI and AD\cite{Taler2008, Boschi2017}. Research therefore should look at these features in more detail and a clarification of this currently disorganised picture should go some way to helping researchers further understand the disease and it's progression. Another area of focus for research of this nature is the process of collecting appropriate language samples. Whilst collecting samples of language is comparatively unintrusive, researchers recognise that these samples require a rich sample of language that potentially cannot be generated by tasks such as the picture description task. Therefore, it would be useful to explore whether spontaneous discourse such a semi-structured interview, has the ability to put pressure on both the cognitive and linguistic systems in the same way as traditional cognitive tests such that it might be able to distinguish between healthy controls, those with MCI and those with AD. There is some evidence to support this. Berisha et al \cite{Berisha2015}, has shown through a longitudinal language analysis of spontaneous speech that there are marked differences in this process between those who would go on to have a diagnosis of AD and a healthy control. 

@@

Alzheimer's Disease (AD) is a neurodegenerative disease in which, from a physiological perspective, the brain develops neurofibrillary tangles and neuritic plaques along with the deterioration and loss of cortical neurons and synapses. Whilst a definitive diagnosis of Dementia can only be produced at post-mortem, there are a number of clinical indicators from psychological perspective that can indicate dementia is present. From a clinical psychology perspective, those who have dementia demonstrate cognitive deficits such as problems with episodic and semantic memory, organizing and planning, difficulties with language and visuospatial deficits\cite{McKhann2011}. In addition, these symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. 

@@

Current thinking suggests that the cognitive deficits associated with AD often begin before the clinical symptoms of the disease become apparent. Researchers propose that neurofibrillary tangles and other associated physiological effects of AD develop over time and alter cognitive function until a threshold is reached and clinical symptoms become more obvious \cite{Nestor2006}. The case of Iris Murdoch, who had a confirmed diagnosis of dementia, illustrates this theory well. Le et al \cite{Le2011}found, in their analysis of three writers and the novels they wrote, that Iris Murdoch's work declined subtly over time, but there was a steep drop off in the use of language in her last novel when, it is theorized, the symptoms of AD manifested themselves more significantly. If this theory holds true more generally, it should be possible to detect subtle cognitive changes in language and memory before a clinical diagnosis can be formed. However, one of the challenges of this approach is differentiating natural cognitive decline due to aging with decline due to a form of dementia. Albert and his team have worked to define clinical criteria which professionals can use to diagnose MCI due to AD and differentiate this from age-associated memory impairment and age-associated cognitive decline. They note that the diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}.

@@

However, whilst the authors claim that the RBANS is adequately sensitive in person's with MCI, other research points out that the RBANS has poor sensitivity in detecting MCI. Another drawback of this battery is the lack of executive function measures and object naming tasks. Research has shown that the RBANS has good test-retest reliability and convergent validity. However, Duff et al warned that caution should be exercised when using the RBANS in a MCI population as it has lower sensitivity in this population \cite{Duff2010}.

@@

Research has consistently shown that MCI patients have a significantly lower digit span score in both normal and reverse order versions of this test and a study by Muangpaisan has shown that the reverse order version of this test can, to some degree, predict a diagnosis of MCI. Muangpaisan also found that Age, Gender and Education have an impact on the performance of the tests. Emrani et al found that the DST revealed no differences between specifically amnestic MCI and controls, but could differentiate between Mixed MCI and the other groups with mixed MCI recalling fewer correct responses than other groups. Notably, there was an attenuated recency effect in those with mixed MCI. \cite{Emrani2018}. Kessels identifies that that there are working memory deficits in MCI patients and these worsen with AD patients. \cite{Kessels2011} He also identified that both MCI and AD have impaired performance on all three conditions of the digit span test. No differences were found between forward and backward conditions in any of the groups. However, available tests may not detect subtle impairments \cite{Kessels2015}.

@@

Rey's Auditory Verbal Learning Test (RAVLT+) looks a wide range of neuropsychological processess including short-term auditory-verbal memory, learning and retention of information. Participants are given a list of 15 unrelated words, repeated over five different trials and are asked to repeat. Another list of 15 unrelated words are given and the client must again repeat the original list of 15 words and then repeat it again after 30 minutes \cite{Rey1941}.

@@

The Rey-Osterrieth Complex Figure (ROCF) is a task widely as a test of visuo-spatial skill and visual memory. The task, which was originally designed by Rey (1944) and standardised by Osterrieth (1944) \cite{Rey1941}, asks a participant to copy a complex geometrical figure (known as the immediate copy condition and to recall and reproduce the figure from memory without warning (known as the delayed recall condition). In the immediate copy condition, the complexity of the figure requires an integrative cognitive ability. The reproduction of such a complex structure involves processes such as planning and organizational strategies that are related to executive functions.

>>>>

In 2009, the Department of Health designed it's National Dementia strategy and as part of this made early diagnosis and support one of it's key priorities \cite{England2009}. A lot of work has gone into trying to find ways of improving the early diagnosis of Alzheimer's Disease (AD) and Mild Cognitive Impairment (MCI) with research focused on two areas - identifying biological markers and analyzing the cognitive decline of those who are suspected to have the disease \cite{Taler2008}. As described above, the numbers of those suffering from AD and MCI are going to increase as the population ages \cite{Prince2015} and thus it is important that we utilize technology wherever possible to aid clinicians in the detection of MCI and AD. At the present time diagnosis is typically conducted at memory clinics by trained clinicians \cite{Boschi2017}. I theorize that we may be able to enable an earlier diagnosis of those with MCI and AD using samples of spontaneous speech, natural language processing (NLP) and machine learning (ML).

@@

There is a large body of research that looks at the decline in language in those with MCI and AD \cite{Taler2008, Boschi2017}. However there is conflicting evidence in these studies about which declining language factors are associated of MCI and AD \cite{Taler2008, Boschi2017}. Research therefore should look at these features in more detail and a clarification of this currently disorganised picture should go some way to helping researchers further understand the disease and it's progression. Another area of focus for research of this nature is the process of collecting appropriate language samples. Whilst collecting samples of language is comparatively unintrusive, researchers recognise that these samples require a rich sample of language that potentially cannot be generated by tasks such as the picture description task. Therefore, it would be useful to explore whether spontaneous discourse such a semi-structured interview, has the ability to put pressure on both the cognitive and linguistic systems in the same way as traditional cognitive tests such that it might be able to distinguish between healthy controls, those with MCI and those with AD. There is some evidence to support this. Berisha et al \cite{Berisha2015}, has shown through a longitudinal language analysis of spontaneous speech that there are marked differences in this process between those who would go on to have a diagnosis of AD and a healthy control. 

@@

Alzheimer's Disease (AD) is a neurodegenerative disease in which, from a physiological perspective, the brain develops neurofibrillary tangles and neuritic plaques along with the deterioration and loss of cortical neurons and synapses. Whilst a definitive diagnosis of Dementia can only be produced at post-mortem, there are a number of clinical indicators from psychological perspective that can indicate dementia is present. From a clinical psychology perspective, those who have dementia demonstrate cognitive deficits such as problems with episodic and semantic memory, organizing and planning, difficulties with language, problems with executive function and visuospatial deficits \cite{McKhann2011}. In addition, these symptoms are often accompanied by emotional problems such as depression and behavioural difficulties. 

@@

Current thinking suggests that the cognitive deficits associated with AD often begin before the clinical symptoms of the disease become apparent. Researchers propose that neurofibrillary tangles and other associated physiological effects of AD develop over time and alter cognitive function until a threshold is reached and clinical symptoms become more obvious \cite{Nestor2006}. The case of Iris Murdoch, who had a confirmed diagnosis of dementia, illustrates this theory well. Le et al \cite{Le2011} found, in their analysis of three writers and the novels they wrote, that Iris Murdoch's work declined subtly over time, but there was a steep drop off in the use of language in her last novel when, it is theorized, the symptoms of AD manifested themselves more significantly. If this theory holds true more generally, it should be possible to detect subtle cognitive changes in language and memory before a clinical diagnosis can be formed. One of the challenges of this approach is differentiating natural cognitive decline due to aging with decline due to a form of dementia. Albert and his team have worked to define clinical criteria which professionals can use to diagnose MCI due to AD and differentiate this from age-associated memory impairment and age-associated cognitive decline. They note that the diagnosis of MCI requires evidence of intra-individual change and optimally requires evaluation at two or more points \cite{Albert2011}.

@@

However, whilst the authors claim that the RBANS is adequately sensitive in person's with MCI \cite{Randolph1998} and there is some research to support this assertion \cite{Karantzoulis2013}, other research points out that the RBANS has poor sensitivity in detecting MCI \cite{Duff2010}. Another drawback of this battery is the lack of executive function measures and object naming tasks. Research has shown that the RBANS has good test-retest reliability and convergent validity. However, Duff et al warned that caution should be exercised when using the RBANS in a MCI population as it has lower sensitivity in this population \cite{Duff2010}.

@@

Research has consistently shown that MCI patients have a significantly lower digit span score in both normal and reverse order versions of this test and a study by Muangpaisan has shown that the reverse order version of this test can, to some degree, predict a diagnosis of MCI \cite{Muangpaisan2008}. Muangpaisan also found that Age, Gender and Education have an impact on the performance of the tests. Emrani et al found that the DST revealed no differences between specifically amnestic MCI and controls, but could differentiate between Mixed MCI and the other groups with mixed MCI recalling fewer correct responses than other groups. Notably, there was an attenuated recency effect in those with mixed MCI. \cite{Emrani2018}. Kessels identifies that that there are working memory deficits in MCI patients and these worsen with AD patients. \cite{Kessels2011} He also identified that both MCI and AD have impaired performance on all three conditions of the digit span test. No differences were found between forward and backward conditions in any of the groups. However, available tests may not detect subtle impairments \cite{Kessels2015}.

@@

Rey's Auditory Verbal Learning Test (RAVLT+) looks a wide range of neuropsychological processess including short-term auditory-verbal memory, learning and retention of information. Participants are given a list of 15 unrelated words, repeated over five different trials and are asked to repeat. Another list of 15 unrelated words are given and the client must again repeat the original list of 15 words and then repeat it again after 30 minutes \cite{Schmidt1996}.

@@

The Rey-Osterrieth Complex Figure (ROCF) is a task widely as a test of visuo-spatial skill and visual memory. The task, which was originally designed by Rey (1944) and standardised by Osterrieth (1944) \cite{Rey1941}, asks a participant to copy a complex geometrical figure (known as the immediate copy condition and to recall and reproduce the figure from memory without warning (known as the delayed recall condition). In the immediate copy condition, the complexity of the figure requires an integrative cognitive ability. The reproduction of such a complex structure involves processes such as planning and organizational strategies that are related to executive functions. In order to make this task repeatable, Taylor et al designed a comparable set of figures which have proven to be of equivalent difficulty \cite{Taylor1969, Strauss1990}. 
